---
title: "dplyr reference"
author: "Yixiang Luo"
date: "`r format(Sys.time(), '%B %d, %Y')`"
bibliography: references.bib
header-includes:
   - \usepackage{float}
   - \usepackage{bbm}
   - \usepackage{amssymb}
   - \usepackage{amsmath}
   - \usepackage{amsthm}
   - \usepackage{mathtools}
   - \usepackage{mathrsfs}
output: 
  pdf_document:
    citation_package: natbib
    number_sections: true
---


```{r setup, echo = FALSE, message=FALSE, warning=FALSE}
# load in useful packages
library(dplyr)



# set default knitr chunks
knitr::opts_chunk$set(
  echo = FALSE,  # don't print the code chunk
  warning = FALSE,  # don't print warnings
  message = FALSE,  # don't print messages
#  fig.width = 6,  # set default width of figures
 fig.height = 4,  # set default height of figures
#  dpi=300, 
  fig.align = "center",  # always align figure in center
  fig.pos = "H",  # always plot figure at the exact location of the code chunk
  cache = FALSE)  # don't cache results

```

``` {r}
# example data
dim(starwars)

slice(starwars, 1:10)
# View(starwars)
```

# dplyr

https://dplyr.tidyverse.org/reference/index.html

## Single table verbs
dplyr aims to provide a function for each basic verb of data manipulation. These verbs can be organised into three categories based on the component of the dataset that they work with:

- Rows:
  - filter() chooses rows based on column values.
  - slice() chooses rows based on location.
  - arrange() changes the order of the rows.
- Columns:
  - select() changes whether or not a column is included.
  - rename() changes the name of columns.
  - mutate() changes the values of columns and creates new columns.
  - relocate() changes the order of the columns.
- Groups of rows:
  - summarise() collapses a group into a single row.

1. filter
- what exactly is the condition?
  treat varname as df$varname and return a vector of T/F
``` {r}
starwars %>% filter(height > 101:187)
# next line is equiv to previous line. Don't be confused with (hair_color == "blond") and think all non-blond are kept, since (hair_color == "blond") is a vector and only the first element is taken by "if"
starwars %>% filter(if(hair_color == "blond") height > 101:187 else T)
starwars %>% filter(rep(T, 87))  # everyone remains
starwars %>% filter(rep(T, 86))  # error of mismatching dim
```

2. slice: pick rows
``` {r}
starwars %>% slice(c(3,1,2))
starwars %>% group_by(hair_color) %>% slice(1)
```

3. arrange(col) = slice(order(df$col))
- NA are always sorted to the end for local data
- grouped arrange ignores groups unless specifically ask: .by_group = TRUE
``` {r}
starwars %>% arrange(height)
starwars %>% arrange(c(3,2,1,4:87))

starwars %>% group_by(hair_color) %>% arrange(height)
starwars %>% group_by(hair_color) %>% arrange(height, .by_group = T)
```

4. select: select column by certain dplyr rules
- `c() for combining selections`
- `: for selecting a range of consecutive variables`
- `! for taking the complement of a set of variables`
- indeces
- `matches(): Matches a regular expression`
``` {r}
starwars %>% select(c("height", "hair_color"))
starwars %>% select("height":"hair_color")
starwars %>% select(1:3)
starwars %>% select(matches("^h.+"))
```

5. mutate(col_name = vector value)
``` {r}
starwars %>% mutate(hello = 1:87) %>% select(matches("^h.+"))
```

6. group_by(vector values (intersection)), ungroup()
``` {r}
starwars %>% group_by(c(rep(0, 3), rep(1, 84))) %>% summarise(max_hight = max(height))
starwars %>% group_by(height_binned = cut(height, 3))

starwars %>% group_by(hair_color, skin_color)
starwars %>% group_by(paste(hair_color, skin_color))

starwars %>% group_by(hair_color, skin_color) %>% ungroup(skin_color)
```
``` {r}
# verbes act on each group
starwars %>% group_by(hair_color) %>% filter(rep(T, 87))   # error
starwars %>% group_by(hair_color) %>% mutate(a = 1:87, .before = "name")   # error
starwars %>% group_by(hair_color) %>% mutate(group_count = length(hair_color), .before = "name")
starwars %>% group_by(hair_color) %>% arrange(height, .by_group = T)   # need .by_group = T
```


7. summarise(): collapse groups
``` {r}
starwars %>% group_by(c(rep(0, 3), rep(1, 84))) %>% summarise(max_hight = max(height), mean_BY = mean(birth_year, na.rm = T))
```

8. reframe(): like summarise, but can return arbitrary lengthed output
``` {r}
starwars %>% group_by(sex) %>% reframe(stat = c(max(height, na.rm = T), median(height, na.rm = T), min(height, na.rm = T)))
```

9. pivot
``` {r}
library(tidyr)
relig_income
long <- relig_income %>% pivot_longer(!"religion", names_to = "income", values_to = "count")
long
long %>% pivot_wider(names_from = income, values_from = count)
```


## table manipulating
1. bind
``` {r}
foo <- data.frame(a = 1:87, b = T)
cbind(foo, starwars)
rbind(data.frame(a = 1, b = "c"), foo)
```

2. set operations using the rows of a data frame.

`intersect(x, y)` finds all rows in both x and y.

`union(x, y)` finds all rows in either x or y, excluding duplicates.

`union_all(x, y)` finds all rows in either x or y, including duplicates.

`setdiff(x, y)` finds all rows in x that aren't in y.

`symdiff(x, y)` computes the symmetric difference, i.e. all rows in x that aren't in y and all rows in y that aren't in x.

`setequal(x, y)` returns TRUE if x and y contain the same rows (ignoring order).

Note that intersect(), union(), setdiff(), and symdiff() remove duplicates in x and y

``` {r}
df1 <- data.frame(a = 1, b = 1:3)
df2 <- data.frame(a = 1, b = 3:5)
df3 <- data.frame(a=1, c=3:5)

# must have the same columns. intersect(df1, df3) -> error
intersect(df1, df2)
union(df1, df2)
union_all(df1, df2)
setdiff(df1, df2)
setdiff(df2, df1)
symdiff(df1, df2)

setequal(df1, df2)
setequal(df2, df3)    # same value, different col names -> False
```

3. join
- inner_join() only keeps observations from x that have a matching key in y.
- left_join() keeps all observations in x.
- right_join() keeps all observations in y.
- full_join() keeps all observations in x and y.
- cross_join() match each row in x to every row in y, resulting in a data frame with nrow(x) * nrow(y) rows. (no `by` argument)

join condition:
`join_by(x$col1 == y$col2, closest(col3 >= col4))`
- simplification: left col is from left df; just one col name if shared name & use ==
- not accept general conditioning, only take `==`, `>=`, `>`, `<=`, `<`, `closest()`, connected by comma ("and" logic). Don't accept expression.
  - what if want expression? e.g. `x$days <= y$days + 1`. mutate an auxiliary column beforehand.
  - what if need more complicated conditioning? Use a more general join (e.g. cross_join() always) and then filter().
- closest(x >= y) is interpreted as: For each value in x, find the closest value in y that is less than or equal to that x value.
closest() will always use the left-hand table (x) as the primary table, and the right-hand table (y) as the one to find the closest match in
``` {r}
sales <- tibble(
  id = c(1, 1, 1, 2, 3),
  sale_date = as.POSIXct(c("2018-12-31", "2019-01-02", "2019-01-05", "2019-01-04", "2019-01-01")),
  other = "a"
)
promos <- tibble(
  id = c(1, 1, 2),
  promo_date = as.POSIXct(c("2019-01-01", "2019-01-05", "2019-01-02"))
)

inner_join(sales, promos, by = "id")

inner_join(sales, promos, by = join_by(id == id, x$sale_date <= y$promo_date))

left_join(sales, promos, by = join_by(id, closest(sale_date <= promo_date)))

cross_join(sales, promos) %>% filter(id.x == id.y | sale_date <= promo_date)
```

## Apply functions
1. across: use with mutate(), mutate a new col for each old col.
`across(.cols, .fns = list of functions, .names = "{.col}_{.fn}")` apply `.fns` to `.cols`.
``` {r}
starwars %>% mutate(across("height":"mass", .fns = list(mean = ~ mean(.x, na.rm = T), shift = ~ .x+1), .names = "{.col}-{.fn}"), .before = "name")
```

2. `if_any` and `if_all`
same as `across`, but expect T/F output and then take a all/any logic.
``` {r}
starwars %>% filter(if_any("height":"mass", .fns = list(cond1 = ~ .x < 70, cond2 = ~ .x > 100)))
starwars %>% mutate(a = if_any("height":"mass", .fns = list(cond1 = ~ .x < 70, cond2 = ~ .x > 100)), .before = "name")
```







